---
title: "Life Expectancy - Linear Model"
author: "Mathieu"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(comment = "")
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
# clean variables
rm(list = ls())
```

## Importing libraries
```{r}
library(tidyverse)

# color palettes
library(hrbrthemes)
library(viridis)

# train-test split
library(rsample)

# tables
library(formattable)
library(huxtable)
library(gt)

# skewness,...
library(moments)
library(patchwork)

# correlation
library(corrplot)

# string manipulation
library(stringr)

# vif test
library(car)

# lasso model selection
library(glmnet)
library(olsrr)

# better summary for linear models
library(broom)

# test for heteroskedasticity
library(lmtest)

# test for normality of the residuals
library(tseries)

# standardized (= studentized) deleted residuals
library(MASS)
```

```{r}
source("utils/plots.r")
source("utils/tables.r")
source("utils/misc.r")
```

## Importing dataset
```{r}
df <- read.csv("dataset/life-expectancy-2012.csv", sep = ",", dec = ".", header = TRUE)

# convert status and adult.mortality to factor (gives an order to categorical variable)
df$status <- factor(df$status, levels = c("Developing", "Developed"))
df$adult.mortality <- factor(df$adult.mortality, levels = c("very low", "low", "middle", "high", "very high"))

df <- df %>%
  dplyr::select(-country)

glimpse(df)
```

## Exploratory Data Analysis

### Qualitative variables

```{r}
par(mfrow = c(1, 2))

table(df$status)

status_percentage <- prop.table(table(df$status))
status_plot <- barplot(status_percentage, col = "#3F6D9B", ylab = "Percentage", main = "Status of countries")

adult_mortality_percentage <- prop.table(table(df$adult.mortality))
adult_mortality_plot <- barplot(adult_mortality_percentage, col = "#3F6D9B", ylab = "Percentage", main = "Adult mortality")
```

#### Quantitative variables

```{r out.width="80%"}
df %>%
  dplyr::select_if(is.numeric) %>%
  summary_moments %>%
  gt %>%
  tab_header(
    title = "4 moments of quantitative variables"
  ) #%>%
  #gtsave("report/figures/quantitative_variable_moments.png", expand = 10)
```

```{r}
df_economics <- df %>%
  dplyr::select(c(percentage.expenditure, total.expenditure, gdp, income.composition.of.resources))

df_socials <- df %>%
  dplyr::select(c(population, schooling))

df_mortality <- df %>%
  dplyr::select(c(infant.deaths, hiv.aids, thinness.5.9.years, thinness.10.19.years))

df_immunization <- df %>%
  dplyr::select(c(hepatitis.b, measles, polio, diphtheria))

df_other <- df %>%
  dplyr::select(c(alcohol, bmi))
```

```{r}
# scatter_plot2 <- function(df, x, y) {
#   ggplot(df, aes({{x}})) +
#     geom_histogram()
# }

# scatter_plot2(df, bmi, gdp)
```

```{r}
gridExtra::grid.arrange(histplot(df, life.expectancy), bplot(df, life.expectancy), nrow = 1)
```

```{r}
gridExtra::grid.arrange(
  histplot(df, percentage.expenditure), 
  histplot(df, total.expenditure),
  histplot(df, gdp),
  histplot(df, income.composition.of.resources),
  nrow = 2, ncol = 2
)

gridExtra::grid.arrange(
  bplot(df, percentage.expenditure), 
  bplot(df, total.expenditure),
  bplot(df, gdp),
  bplot(df, income.composition.of.resources),
  nrow = 2, ncol = 2
)
```


```{r}
gridExtra::grid.arrange(
  histplot(df, population), 
  histplot(df, schooling),
  bplot(df, population), 
  bplot(df, schooling),
  nrow = 2, ncol = 2
)
```

```{r}
gridExtra::grid.arrange(
  histplot(df, infant.deaths),
  histplot(df, hiv.aids),
  histplot(df, thinness.5.9.years),
  histplot(df, thinness.10.19.years),
  nrow = 2, ncol = 2
)

gridExtra::grid.arrange(
  bplot(df, infant.deaths),
  bplot(df, hiv.aids),
  bplot(df, thinness.5.9.years),
  bplot(df, thinness.10.19.years),
  nrow = 2, ncol = 2
)
```

```{r}
gridExtra::grid.arrange(
  histplot(df, hepatitis.b), 
  histplot(df, measles),
  histplot(df, polio),
  histplot(df, diphtheria),
  nrow = 2, ncol = 2
)

gridExtra::grid.arrange(
  bplot(df, hepatitis.b), 
  bplot(df, measles),
  bplot(df, polio),
  bplot(df, diphtheria),
  nrow = 2, ncol = 2
)
```

```{r}
gridExtra::grid.arrange(
  histplot(df, alcohol), 
  histplot(df, bmi),
  bplot(df, alcohol), 
  bplot(df, bmi),
  nrow = 2, ncol = 2
)
```



### Correlation matrix

```{r}
corr_matrix <- cor(keep(df, is.numeric))
corrplot(
  corr_matrix, 
  method = "shade", 
  shade.col = NA, 
  number.cex = 0.3,
  tl.col = "black", 
  tl.src = 90,
  tl.cex = 0.7,
  addCoef.col = "black", 
  order = "hclust"
)
```
Let's do some cleaning after interpreting our correlation plot
```{r}
df <- df %>%
  dplyr::select(-under.five.deaths)
```

### Model selection

We convert categorical variables into dummies variables (one-hot-encoding)
```{r}
df_one_hot <- df
df_one_hot <- one_hot_encoding(df_one_hot, c("status", "adult.mortality"))

head(df_one_hot)
```

```{r}
corr_matrix <- cor(keep(df_one_hot, is.numeric))
corrplot(
  corr_matrix, 
  method = "shade", 
  shade.col = NA, 
  number.cex = 0.3,
  tl.col = "black", 
  tl.src = 90,
  tl.cex = 0.7,
  addCoef.col = "black", 
  order = "hclust"
)
```


Let's split dataset into 80-20
```{r}
set.seed(123)

split <- initial_split(df_one_hot, prop = 0.8)
df_train <- training(split)
df_test <- testing(split)

head(df_train)
```

```{r}
#df_train <- df_train %>%
#  dplyr::select(-c(gdp, income.composition.of.resources))

#df_test <- df_test %>%
#  dplyr::select(-c(gdp, income.composition.of.resources))
```

```{r}
full_lm <- lm(life.expectancy ~ ., data = df_train)
intercept_only <- lm(life.expectancy ~ 1, data = df_train)
```

#### Cp Mallow
```{r}

``` 

#### Forward stepwise selection

```{r}
# direction = "forward" => forward selection ; direction = "backward" => backward selection 
# direction = "both" => forward or backward stepwise selection (forward and backward is chosen with scope parameter)

# k = 2 => use AIC criterion (tendancy too select too much variables and overfit)
forward_stepwise_aic_lm <- step(
  intercept_only,
  direction = "both", 
  scope = list(lower = life.expectancy ~ 1, upper = life.expectancy ~ .),
  k = 2
)

backward_stepwise_aic_lm <- step(
  full_lm, 
  direction = "both", 
  scope = list(lower = life.expectancy ~ 1, upper = life.expectancy ~ .), 
  k = 2,
  trace = 0
)

# k = log(n) => use BIC criterion (penalize more but can underfit)
forward_stepwise_bic_lm <- step(
  intercept_only,
  direction = "both", 
  scope = list(lower = life.expectancy ~ 1, upper = life.expectancy ~ .), 
  k = log(nrow(df_train)),
  trace = 0
)

backward_stepwise_bic_lm <- step(
  full_lm, 
  direction = "both", 
  scope = list(lower = life.expectancy ~ 1, upper = life.expectancy ~ .), 
  k = log(nrow(df_train)),
  trace = 0
)

full_model_summary <- model_comparison_table(full_lm, "Full model")
forward_stepwise_aic_summary <- model_comparison_table(forward_stepwise_aic_lm, "Forward stepwise AIC")
backward_stepwise_aic_summary <- model_comparison_table(backward_stepwise_aic_lm, "Backward stepwise AIC")
forward_stepwise_aic_summary <- model_comparison_table(forward_stepwise_aic_lm, "Forward stepwise AIC")
backward_stepwise_bic_summary <- model_comparison_table(backward_stepwise_bic_lm, "Backward stepwise BIC")
```

#### LASSO 

```{r}
response <- df_train$life.expectancy
explanatory <- df_train %>% 
  dplyr::select(-life.expectancy)

# alpha = 1 => LASSO (note: alpha = 0 => ridge regression)
# perform k-fold with k = 10 to select the best lambda (that minimises MSE)
cv_model <- cv.glmnet(x = as.matrix(explanatory), y = response, alpha = 1)

best_lambda <- cv_model$lambda.min
lasso <- glmnet(x = explanatory, y = response, alpha = 1, lambda = best_lambda)

lasso_formula <- get_model_formula(lasso)

lasso_lm <- lm(lasso_formula, data = df_train)
lasso_summary <- model_comparison_table(lasso_lm, "LASSO")
```

```{r}
model_selection_table <- rbind(full_model_summary, forward_stepwise_aic_summary, backward_stepwise_aic_summary, forward_stepwise_aic_summary, backward_stepwise_bic_summary, lasso_summary)
gt(model_selection_table)
```

#### Interactions with variables

```{r}
chosen_model_formula <- get_model_formula(backward_stepwise_aic_lm)

model1 <- lm(paste(chosen_model_formula, "+ infant.deaths * measles"), data = df_train)
model2 <- lm(paste(chosen_model_formula, "+ adult.mortality.high * alcohol"), data = df_train)
model3 <- lm(paste(chosen_model_formula, "+ adult.mortality.very_high * alcohol"), data = df_train)
model4 <- lm(paste(chosen_model_formula, "+ total.expenditure * adult.mortality.low"), data = df_train)
model5 <- lm(paste(chosen_model_formula, "+ total.expenditure * infant.deaths"), data = df_train)
model6 <- lm(paste(chosen_model_formula, "+ percentage.expenditure * diphtheria"), data = df_train)

interaction_table <- rbind(
  backward_stepwise_aic_summary,
  model_comparison_table(model1, "infant.deaths * measles"),
  model_comparison_table(model2, "adult.mortality.high * alcohol"),
  model_comparison_table(model3, "adult.mortality.very_high * alcohol"),
  model_comparison_table(model4, "total.expenditure * adult.mortality.low"),
  model_comparison_table(model5, "total.expenditure * infant.deaths"),
  model_comparison_table(model6, "percentage.expenditure * diphtheria")
)

gt(interaction_table)
```

```{r}
final_lm <- model5
```

#### Nonlinearity
```{r}
final_lm_variables <- tidy(final_lm)$term
print(final_lm_variables[0:-1])



gridExtra::grid.arrange(
  scplot_residuals_variable(final_lm, total.expenditure),
  scplot_residuals_variable(final_lm, hiv.aids),
  scplot_residuals_variable(final_lm, income.composition.of.resources),
  scplot_residuals_variable(final_lm, adult.mortality.low),
  scplot_residuals_variable(final_lm, adult.mortality.middle),
  scplot_residuals_variable(final_lm, adult.mortality.very_high),
  scplot_residuals_variable(final_lm, infant.deaths),
  ncol = 3
)
```

#### Outliers and influential observations
```{r}
n <- nrow(df_train)
p <- length(final_lm$coefficients) - 1

# outliers with respect to the explanatory variables

## outliers 
# diagonal of the hat matrix
h_ii <- lm.influence(final_lm)$hat
outliers_criterion <- 2 * p / n
outliers_criterion

which(h_ii > outliers_criterion)
length(which(h_ii > outliers_criterion))

## influentials on fitted values
dffits <- dffits(final_lm)
influential_fitted_criterion <- 2 * sqrt(p / n)
influential_fitted_criterion

which(abs(dffits) > influential_fitted_criterion)
length(which(abs(dffits) > influential_fitted_criterion))

plot(dffits, type = "h", xlab = "observation")
abline(h = influential_fitted_criterion, lty = 2)
abline(h = -influential_fitted_criterion, lty = 2)

## influentials on regression coefficients
cooks <- cooks.distance(final_lm)
plot(cooks, xlab = "observations")
abline(h = 4/n, lty = 2, col = "steelblue")

# outliers with respect to the response variable 
di_ast <- studres(final_lm)
student_quantile <- qt(0.975, df = n - p - 1)
which(di_ast > student_quantile)
```

#### Multicollinearity
```{r}
var_inf_factor = ols_vif_tol(final_lm)
var_inf_factor_table = within(var_inf_factor, {
  Tolerance = round(Tolerance, 2)
  VIF = round(VIF, 2)
})

round(mean(var_inf_factor$VIF), 2)

gt(var_inf_factor_table) %>%
  tab_header(title = "Variation inflation factor (VIF)", subtitle = "final model") #%>%
  #gtsave("report/figures/vif_final_model.png")
```

```{r}

response <- df_train$life.expectancy

explanatory_variables <- str_replace(tidy(final_lm)$term[0:-1], ":", "_")
explanatory_matrix <- df_train %>% 
  mutate(total.expenditure_infant.deaths = total.expenditure * infant.deaths) %>%
  dplyr::select(explanatory_variables) %>%
  as.matrix

df_train_with_interaction_term <- df_train %>%
  mutate(total.expenditure_infant.deaths = total.expenditure * infant.deaths) 

# k-fold cross-validation to find optimal lambda value (kappa in lecture notes)
cv_model <- cv.glmnet(x = explanatory_matrix, y = response, alpha = 0)
#plot(cv_model)

best_lambda <- cv_model$lambda.min
ridge <- glmnet(x = explanatory_matrix, y = response, alpha = 0, lambda = best_lambda)

ridge_formula <- get_model_formula(ridge)

ridge_lm <- lm(ridge_formula, data = df_train_with_interaction_term)
ridge_summary <- model_comparison_table(ridge_lm, "RIDGE")

gt(ridge_summary) %>%
  tab_header(title = "Ridge regression applied on selected model")
```
Let's check if multicollinearity is fixed
```{r}
var_inf_factor = ols_vif_tol(ridge_lm)
var_inf_factor_table = within(var_inf_factor, {
  Tolerance = round(Tolerance, 2)
  VIF = round(VIF, 2)
})

round(mean(var_inf_factor$VIF), 2)

gt(var_inf_factor_table) %>%
  tab_header(title = "Variation inflation factor (VIF)", subtitle = "after correction with ridge") #%>%
  #gtsave("report/figures/vif_final_model.png")
```

#### Heteroskedasticity

```{r}
final_lm_variables <- tidy(final_lm)$term
print(final_lm_variables[0:-1])

gridExtra::grid.arrange(
  scplot_residuals_fitted(final_lm, total.expenditure),
  scplot_residuals_fitted(final_lm, hiv.aids),
  scplot_residuals_fitted(final_lm, income.composition.of.resources),
  scplot_residuals_fitted(final_lm, adult.mortality.low),
  scplot_residuals_fitted(final_lm, adult.mortality.middle),
  scplot_residuals_fitted(final_lm, adult.mortality.very_high),
  scplot_residuals_fitted(final_lm, infant.deaths),
  ncol = 3
)
```

```{r}
bptest(final_lm, data = df_train)
```

```{r}
h1 <- histplot(df_train, life.expectancy)
h2 <- histplot(df_train, log(life.expectancy))

gridExtra::grid.arrange(h1, h2, ncol = 2)
```

```{r}
df_train_final_model <- df_train_with_interaction_term %>%
  dplyr::select(c(life.expectancy, explanatory_variables))

head(df_train_final_model)

final_lm_log <- lm(log(life.expectancy) ~ ., data = df_train_final_model)
```

```{r}
gridExtra::grid.arrange(
  scplot_residuals_fitted(final_lm_log, total.expenditure),
  scplot_residuals_fitted(final_lm_log, hiv.aids),
  scplot_residuals_fitted(final_lm_log, income.composition.of.resources),
  scplot_residuals_fitted(final_lm_log, adult.mortality.low),
  scplot_residuals_fitted(final_lm_log, adult.mortality.middle),
  scplot_residuals_fitted(final_lm_log, adult.mortality.very_high),
  scplot_residuals_fitted(final_lm_log, infant.deaths),
  ncol = 3
)
```

```{r}
bptest(final_lm_log, data = df_train_final_model)
```

#### Autocorrelation
```{r}
bgtest(final_lm, order = 8)
```

#### Normality of the residuals
```{r}
qqnorm(final_lm$residuals, pch = 1, frame = FALSE)
qqline(final_lm$residuals, lwd = 2)

jarque.bera.test(final_lm$residuals)
```